---
title: '6\. Worksheet: Among Site (Beta) Diversity -- Part 2'
author: "Anna Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

In this worksheet, we continue to explore concepts, statistics, and visualizations related to $\beta$-diversity.
Now that you know how to formally quantify $\beta$-diversity, we will learn how to test hypotheses about $\beta$-diversity using multivariate statistics. 

## Directions:
1. In the Markdown version of this document in your cloned repo, change "Student Name" on line 3 (above) with your name.
2. Complete as much of the worksheet as possible during class.
3. Use the handout as a guide; it contains a more complete description of data sets along with examples of proper scripting needed to carry out the exercises.
4. Answer questions in the worksheet.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio (color may vary if you changed the editor theme). 
5. Before you leave the classroom today, you should **push** this file to your GitHub repo, at whatever stage you are. This will enable you to pull your work onto your own computer.
6. When you have completed the worksheet, **Knit** the text and code into a single PDF file by pressing the `Knit` button in the RStudio scripting panel.
This will save the PDF output in your Posit.cloud workspace: `/cloud/project/QB-2025/Week4-Beta/`
7. After Knitting, please submit the worksheet by making a **push** to your GitHub repo and then create a **pull request** via GitHub.
Your pull request should include this file (**6.BetaDiversity_2_Worksheet.Rmd**) 
with all code blocks filled out and questions answered) and the PDF output of `Knitr`  
(**6.BetaDiversity_2_Worksheet.pdf**).

The completed exercise is due on **Wednesday, February 12^th^, 2025 before 12:00 PM (noon)**.

## 1) R SETUP

Typically, the first thing you will do in either an R script or an RMarkdown file is setup your environment. 
This includes things such as setting the working directory and loading any packages that you will need.

In the R code chunk below, provide the code to:  

1. clear your R environment,
2. print your current working directory,
3. set your working directory to your `Week4-Beta/` folder. 
4. load the `vegan` R package (be sure to install if needed).  

```{r}
rm(list=ls())
getwd()
setwd("/cloud/project/QB2025_ALennon/Week4-Beta")
```

## 2) LOADING DATA
### Load dataset
In the R code chunk below, load the `doubs` dataset from the `ade4` package

```{r}
# note, pleae do not print the dataset when submitting

library("lattice")
library("vegan")
library("ade4")
library("viridis")
library("gplots")
library("indicspecies")
library("tidyverse")
library("indicspecies")
library("permute")
#install.packages("virdisLite")
data(doubs)
#print(doubs)

```
## 3) HYPOTHESIS TESTING

### A.  Multivariate Procedures for Categorical Designs
Earlier work done in the Doubs River suggested that the river has four distinct regions of habitat quality: the first region (sites 1-14) of "high quality"; the second (sites 15 - 19) and fourth (sites 26 - 30) of "moderate quality"; and the third  (sites 20 - 25) of "low quality".

In the code chunk below, test the hypothesis that fish community composition varies with river quality.

1. create a factor vector that categorizes habitat quality in the Doubs River,
2. use the multivariate analyses for categorical predictors to describe how fish community structure relates to habitat quality.

```{r}
fish <-doubs$fish
fish <-fish[-8,]
quality <- c(rep("HQ", 13), rep("MQ", 5), rep("LQ", 6), rep("MQ", 5))
#print(quality)
#PERMANOVA
adonis2(fish ~ quality, method = "bray", permutations = 999)

#Species-Site 
indval <- multipatt(fish, cluster = quality, func = "IndVal.g", 
          control = how(nperm=999)) 
summary(indval)

#phi coefficient of association
fish.rel <- decostand(fish, method = "total")
fish.phi <- multipatt(fish.rel, cluster = quality, func = "r.g", 
        control = how(nperm=999)) 
summary(fish.phi)

```

***Question 1***: Based on the PERMANOVA, IndVal, and phi coefficient analyses, what did you learn about the relationship between habitat quality and the fish species composition?
Are the different analyses consistent with one another and do they agree with the visualizations (heat maps, cluster dendograms, ordinations) that you created?

> ***Answer 1***: Overall, the various hypothesis tests are in consensus that habitat quality and fish species composition are highly related at a signficant level. In particular, the high quality habitat are highly correlated with fish species. The results of the hypothesis testing matches with the visualizations. 


### B.  Multivariate Procedures for Continuous Designs

### i.  Mantel Test
In the R code chunk below, do the following:   

1. create distance matrices for both fish communities and environmental factors, and
2. use a Mantel test to test the hypothesis that fish assemblages are correlated with stream environmental variables.

```{r}

fish.dist <- vegdist(doubs$fish[-8, ], method = "bray")
env.dist <- vegdist(scale(doubs$env[-8,]),method = "euclid")

#Mantel Test
mantel(fish.dist,env.dist)
```

***Question 2***: What do the results from our Mantel test suggest about fish diversity and stream environmental conditions? 
How does this relate to your hypothesis about stream quality influencing fish communities?

> ***Answer 2***:  Given the r-statistic is .604 with a p-value of .001, fish diversity and stream environmental conditions are strongly related. Therefore, environmental conditions provide a strong indicator of fish diversity. I had hypothesized that stream quality would impact fish communities and predicted this would be a positive correlation. 

### ii.  Constrained Ordination
In the R code chunk below, do the following:  

1. create an environmental matrix of the water chemistry data included in the `doubs` dataset using forward and reverse selection of variables,
2. conduct a redundancy analysis on the fish assemblages of the Doubs River,
3. use a permutation test to determine the significance of the constrained analysis,
4. use a permutation test to determine the correlation of each environmental factor on the constrained axes,
5. calculate the explained variation on the first and second constrained axes,
6. plot the constrained ordination results including labeled points for each site, and
7. add vectors that demonstrate the influence of each environmental factor the constrained ordination.   
```{r}
#Bray-curtis
fish.db <- vegdist(fish, method = "bray")
fish.db <- vegdist(fish, method = "bray", upper = TRUE, diag = TRUE)
#matrix
env.chem <- as.matrix(doubs$env[-8 , 5:11])
doubs.dbrda <- dbrda(fish.db ~ ., as.data.frame(env.chem))
ordiplot(doubs.dbrda)

#Intercept
doubs.dbrda.mod0 <- dbrda(fish.db ~ 1, as.data.frame(env.chem))
ordiplot(doubs.dbrda.mod0) 

# Full model
doubs.dbrda.mod1 <- dbrda(fish.db ~ ., as.data.frame(env.chem))

#AIC values
doubs.dbrda <- ordiR2step(doubs.dbrda.mod0, doubs.dbrda.mod1, perm.max = 200)

#visual
doubs.dbrda$call
doubs.dbrda$anova
ordiplot(doubs.dbrda)

# Permutation tests
permutest(doubs.dbrda, permutations = 999)
envfit(doubs.dbrda, env.chem[,c(4,6,7)], perm = 999)

# Explained Variation
dbrda.explainvar1 <- round(doubs.dbrda$CCA$eig[1] / 
                    sum(c(doubs.dbrda$CCA$eig, doubs.dbrda$CA$eig)), 3) * 100
dbrda.explainvar2 <- round(doubs.dbrda$CCA$eig[2] / 
                    sum(c(doubs.dbrda$CCA$eig, doubs.dbrda$CA$eig)), 3) * 100
#Plot
par(mar = c(5, 5, 4, 4) + 0.1)

plot(scores(doubs.dbrda, display = "wa"), xlim = c(-1.3, 1.1), 
     ylim = c(-1.1, 2.7), xlab = paste("dbRDA 1 (", dbrda.explainvar1, "%)", 
     sep = ""), ylab = paste("dbRDA 2 (", dbrda.explainvar2, "%)", sep = ""),
     pch = 16, cex = 2.0, type = "n", cex.lab = 1.5, 
     cex.axis = 1.2, axes = FALSE)

axis(side = 1, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
axis(side = 2, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
abline(h = 0, v = 0, lty = 3)
box(lwd = 2)

points(scores(doubs.dbrda, display = "wa"),
       pch = 19, cex = 3, bg = "gray", col = "gray")
text(scores(doubs.dbrda, display = "wa"), 
     labels = row.names(scores(doubs.dbrda, display = "wa")))

# Add  Vectors
vectors <- scores(doubs.dbrda, display = "bp")

arrows(0, 0, vectors[,1], vectors[, 2], 
       lwd = 2, lty = 1, length = 0.2, col = "cadetblue")
text(vectors[,1], vectors[, 2], pos = 3, 
     labels = row.names(vectors))
axis(side = 3, lwd.ticks=2, cex.axis=1.2, las = 1, col = "cadetblue", lwd = 2.2,
     at = pretty(range(vectors[, 1])) * 2, labels = pretty(range(vectors[, 1])))
axis(side = 4, lwd.ticks=2, cex.axis=1.2, las = 1, col = "cadetblue", lwd = 2.2,
     at = pretty(range(vectors[, 2])) * 2, labels = pretty(range(vectors[, 2])))

```

***Question 3***: Based on the constrained ordination, what are the environmental variables (or groups of correlated variables) that seem to be contributing to variation in fish community structure?

> ***Answer 3***: Oxygen ("oxy"), dissolved oxygen content ("bdo"), and nitrogen ("nit") are the three environmental variables that contribute the greatest to variation in the fish community. 

### iii. Variation Partitioning
In the code chunk below,   

1. Create a matrix model of the selected environmental variables,
2. Create a matrix model of the selected PCNM axes,
3. Perform constrained and partial constrained ordinations using the spatial and environmental models you just created,
4. Test the significance of each of your constrained ordinations using permutation tests,
5. Partition the variation among sites into the relative importance of space, environment, spatially structured environment, and residuals,
6. Plot the variation partitioning output to visualize it.  
```{r}
doubs.dbrda$anova
env.mod <- model.matrix(~ oxy + bdo + nit, as.data.frame(env.chem))[,-1]

rs <- rowSums(fish)/sum(fish)

# PCNM
doubs.pcnmw <- pcnm(dist(doubs$xy[-8,]), w = rs, dist.ret = T) 
doubs.pcnmw$values > 0
doubs.space <- as.data.frame(scores(doubs.pcnmw))
doubs.pcnm.mod0 <- dbrda(fish.db ~ 1, doubs.space)
doubs.pcnm.mod1 <- dbrda(fish.db ~ ., doubs.space)
step.pcnm <- ordiR2step(doubs.pcnm.mod0, doubs.pcnm.mod1, perm.max = 200)

plot(step.pcnm)

step.pcnm$anova

space.mod <- model.matrix(~ PCNM2 + PCNM3 + PCNM5 + PCNM1 + 
                            PCNM13 + PCNM16 + PCNM6, doubs.space)[,-1]

doubs.total.env <- dbrda(fish.db ~ env.mod)
doubs.total.space <- dbrda(fish.db ~ space.mod)

# partial constrained ordinations
doubs.env.cond.space <- dbrda(fish.db ~ env.mod + Condition(space.mod)) 
doubs.space.cond.env <- dbrda(fish.db ~ space.mod + Condition(env.mod)) 

# significance dbRDA fractions.
permutest(doubs.env.cond.space, permutations = 999)
permutest(doubs.space.cond.env, permutations = 999)
permutest(doubs.total.env, permutations = 999)
permutest(doubs.total.space, permutations = 999)

# fractions of variation
doubs.varpart <- varpart(fish.db, env.mod, space.mod) 
doubs.varpart

par(mar = c(2,2,2,2))
plot(doubs.varpart)
text(1, 0.25, "Space")
text(0, 0.25, "Env")
mtext("Variation Partitioning of\nDoubs Fish Diversity", side = 3, line = -3)

```

***Question 4***: Interpret the variation partitioning results.   

> ***Answer 4***: For the fish diversity in the Doubs river, .13 of the variation can be accounted from the environment and .24 from Space, meaning Space contributes to more of the variation. However, there is .37 of the variation that is due to a mix of both the environment and space and .27 of the variation that is explained by neither the environment and space. Given the high percentage of variation explained by space, the environment, and both, space and environment are good predictors of Doubs fish variation. 

## SYNTHESIS

Load the dataset from that you and your partner are using for the team project. 
Use one of the hypothesis-testing tools introduced in the beta diversity module. 
Interpret the findings of your data with respect to principles of biodiversity. 

>Running the follwoing code will not allow it to knit in reasonable time- I will upload the code in an R markdown when it finishes running (Anna 11:41 2/12/2025)
setwd("/cloud/project/QB_biodiversity_project_EH")
# Load necessary libraries
library(vegan)
library(tibble)
library(lattice)
library(readxl)
library(dplyr)
library(tidyr)
#load in data from separate R script file
tree <-read.csv("TREE.csv")

>tree.species.df <-data.frame(Plot_ID = c(tree$PLOT),
                             SPCD =c(tree$SPCD))
myco <-as.data.frame(read_xlsx("MycoType_ref2.xlsx"))
myco<- myco[,-c(3:7)]
invasive <-read_xlsx("INVASIVE.xlsx")
untree.species <- unique(tree.species.df$SPCD)
unmyco.species <-unique(myco$SPCD)
tree.species.df <- tree.species.df %>%
  filter(!SPCD %in% c(999, 998)) 

>#add myco type
tree.species.df$PlotStatus <- NA

># Loop through rows of tree.species.df
>for (i in seq_len(nrow(tree.species.df))) {  
  >if (tree.species.df$Plot_ID[i] %in% invasive$PLOT) {  
    >tree.species.df$PlotStatus[i] <- "invasive"} else {  
    >tree.species.df$PlotStatus[i] <- "non-invasive"}}
    
>#Site by Speccies
tree.species.df.1 <-data.frame(Plot_ID = c(tree$PLOT),
                               Species_ID =c(tree$SPCD))
#print(tree.species.df)


>tree.ss.df <- as.data.frame.matrix(table(tree.species.df.1$Plot_ID, tree.species.df.1$Species_ID))
tree.ss.df <- rownames_to_column(tree.ss.df, var = "Plot_ID")
#print(tree.ss.df)

>tree.species.only.df <- tree.ss.df[, 2:ncol(tree.ss.df)]


>#Create Factors vector

>factor <- factor(tree.species.df$PlotStatus)

>length(factor)
nrow(tree.species.only.df)

>factor_trimmed <- factor[1:nrow(tree.species.only.df)]

>adonis2(tree.species.only.df ~ factor_trimmed, method = "bray", permutations = 999)
